More on the Slavnov products of spin chains, KP hierarchy tau and Baker-Akhiezer functions

### 2025-05-12 draft.tex
fix small mistakes
write abstract, intro and improve section 2

\title{Slavnov products and KP tau functions}

@@

\address{\noindent 
Instituto de Ciências Exatas, Departamento de Física\\
Universidade Federal Fluminense\\
Rua Des. Ellis Hermydio Figueira, 783, Aterrado\\
27213-145 Volta Redonda, RJ, Brazil
}

@@

\begin{abstract}
One important ingredient in the Algebraic Bethe Ansatz is the scalar
product of Bethe states.  There are many important results on this
topic and references therein, but here we turn our attention to the
well known fact that scalars products in the algebraic Bethe ansatz
can be written as determinants.  More specifically, we will
investigate a particular type of correlation functions in quantum
integrable spin chains, the so called \emph{Slavnov product} These
objects can be seen as building blocks of more sofisticated
correlation functions and are instrumental in the study of norms of
Bethe wave functions But more importantly for our present discussion,
Slavnov products seem to be a bridge connecting quantum and classical
integrable system.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

One important ingredient in the Algebraic Bethe Ansatz is the scalar
product of Bethe states.  There are many important results on this
topic, see~\cite{Korepin:1993kvr} and references therein, but here we
turn our attention to the well known fact that scalars products in
the algebraic Bethe ansatz can be written as determinants.

More specifically, we will investigate a particular type of
correlation functions in quantum integrable spin chains, the so called
\emph{Slavnov product}~\cite{Slavnov1989}. These objects can be seen as building blocks
of more sofisticated correlation functions and are instrumental in the
study of norms of Bethe wave functions~\cite{Korepin:1993kvr}.  But
more importantly for our present discussion, Slavnov products seem to
be a bridge connecting quantum and classical integrable system.

The study of connections between classical and quantum integrable
systems is an old subject, and has been addressed from a wide range of
perspectives~\cite{Wu:1975mw, Its:1992bj, korepin2000the, Foda2009,
  Alexandrov:2011aa}.  In~\cite{Foda:2009zz}, the authors started the
analysis of the connections between the Slavnov products of the
Heisenberg XXZ spin chain, and the Kadomtsev–Petviashvili (KP)
classical integrable hierarchy. This analysis has been extended to
some directions and applications, for example~\cite{Wheeler:2010vmq,
  Foda:2010, Takasaki:2010qm, Foda:2012wn, Foda:2012wf}.
In~\cite{Araujo:2021ghu, Araujo:2024klz}, the author of the current
paper also investigated some developments of this research field. In
particular, the relation between integrable hierarchies and quantum
integrable systems such as the Q-Boson integrable system and in the
Temperley-Lieb open spin chains.

Although the specific details of he Slavnov product are radically
different, In all these cases discussed in the previous paragraph, the
general functional structure of the Slavnov products of different
models are basically the same. More specifically, module some
multiplicative terms, all these expressions are determinant, and it is
the fact that allows us to manipulate these objects to show that thesy
are KP and Toda tau functions.

Belliard and Slavnov~\cite{Belliard:2019bfz} have addressed precisely
this property: \emph{Why scalar products in the algebraic Bethe ansatz
have determinant representations}.  Using their results, we now want
to understand \emph{why and how (some) scalar products in the
algebraic Bethe ansatz are KP and Toda tau functions}. In the current
show that as long as the integrable system satisfies the conditions
established in~\cite{Belliard:2019bfz} -- and some additional
restrictions -- the Slavnov product is guaranteed to satisfy the KP
integrable hiearchy. Moreover, we also discuss some immediate
consequences of this result.

In section 2 we review the main aspects of the
work~\cite{Belliard:2019bfz} and fix the main notation we use in the
remainder of this work. (...)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Scalar products as determinants}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In order to make the paper as self-contained as possible, this section
reviews the main arguments in~\cite{Belliard:2019bfz}. Let us start
with a set of arbitrary complex numbers \(\bm{u} =
\{u_j\}_{j=1}^{n+1}\), and we set \(n+1\) sets
\(\bm{u}_j=\bm{u}\setminus \{u_j\}\). In the algebraic Bethe ansatz
setting~\cite{Korepin:1993kvr}, we can now define \(n+1\) off-shell
Bethe vectors \(\ket{\Psi(\bm{u}_j)}\), i.e. Bethe states where the
algebraic Bethe equations are not imposed on
\(\bm{u}_j\). Additionally, we also consider a set of Bethe roots
\(\bm{v} = \{ v_k \}_{k=1}^n\), and the on-shell Bethe vectors
\(\ket{\Psi(\bm{v})}\).

\subsection{Determinant representation}
The action of the transfer matrix \(\mathcal{T}(z)\), that is an
Hermitian operator, on the dual Bethe vector is given by
\begin{equation}
\label{eq:eigenvalue-bethe}
\bra{\Psi(\bm{v})}\mathcal{T}(z) = \Lambda(z; \bm{v}) \bra{\Psi(\bm{v})} \; , 
\end{equation}
where \(\Lambda(z; \bm{v}) \) is the eigenvalue of the transfer matrix.

We now introduce \(n+1\) scalar products between the on-shell and
off-shell Bethe states
\begin{equation}
  \bm{\zeta}_j(\bm{u}_j, \bm{v}) \equiv  \bm{\zeta}_j
  = \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)}\qquad j =1, \dots, n+1\; .
\end{equation}
These partially on-shell scalar products are called \emph{Slavnov
products}.

Before procceding, let us also establish the class of models we are
dealing with.  Let us consider a class of integrable models where the
action of the transfer matrix on a generic off-shell bethe states can
be expanded as
\begin{subequations}
\begin{equation}
\label{eq:scalar-exp}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)}  = \sum_{k=1}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; ,
\end{equation}
where \(L_{jk}\) are numerical coefficients. Observe that the terms
\(L_{jk}\) with \(j\neq k\) contain the unwanted terms.; and these
non-diagonal terms must vanish when the Bethe equations are
satisfied. Therefore, since the diagonal coefficient survives, that is
\(L_{jj}\), we conclude that these terms must be equal to the
eigenvalue of the tranfer matrix, that is
\begin{equation}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} =  \Lambda(u_j; \bm{u}_j) \ket{\Psi(\bm{v})} + 
  \sum_{\substack{k=1\\ k\neq j}}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; . 
\end{equation}
\end{subequations}
This class of models include a wide range of known integrable systems,
including the ones discussed in the introduction. One important class
of models that does not fall in this classification are with elliptic
R-matrices, although some of these models have slavnov products with
determinantal representations.

\begin{proposition}
The Slavnov products \(\bm{\zeta}_j\) satisfy a system of linear
equations.
\end{proposition}

\paragraph{\textbf{\emph{Proof}}} Let us first use that
\begin{equation}
\label{eq:scalar-prod}
  \bra{\Psi(\bm{v})} \left(\mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} \right) =
  \left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)\ket{\Psi(\bm{u}_j)}
  \qquad j =1, \dots, n+1\; .
\end{equation}
The right-hand side of this equation can be simplified with the eigenvalue
expression~(\ref{eq:eigenvalue-bethe}), that is
\begin{equation}
  \left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)\ket{\Psi(\bm{u}_j)}  =
\Lambda(u_j; \bm{v}) \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)} = 
\Lambda(u_j; \bm{v}) \bm{\zeta}_j\; . 
\end{equation}

@@

Puttting all these facts together, we get
\begin{equation}
  \sum_{k=1}^{n+1} L_{jk} \bm{\zeta}_k  = 
  \Lambda(u_j; \bar{v}) \bm{\zeta}_j\quad \Rightarrow \quad
  \sum_{k=1}^{n+1} M_{jk} \bm{\zeta}_k = 0 \; ,
\end{equation}
where
\begin{equation}
  M_{jk} = L_{jk} - \delta_{jk}\Lambda(u_j; \bm{v})\; . 
\end{equation}
And this expression shows that the Slavnov products satisfy a linear system.
We can write this system in a matrix form as 
\begin{equation}
\bm{M} \bm{\zeta} = 
  \left(\bm{L} - \bm{\Lambda} \right) \bm{\zeta} = \bm{0}\; ,
\end{equation}
where
\begin{subequations}
  \begin{equation}
    \bm{L} = 
 \begin{pmatrix}
   L_{1,1} & \cdots & L_{1,n+1}\\
   L_{2,1} & \cdots & L_{2,n+1}\\
   \vdots & \vdots & \vdots\\
   L_{n+1, 1} & \cdots & L_{n+1, n+1}\\
 \end{pmatrix}  \quad , \quad
    \bm{\Lambda} = \mathrm{diag}(\Lambda_1, \Lambda_2, \dots, \Lambda_{n+1})
\end{equation}
with \(\Lambda_j \equiv \Lambda(u_j; \bm{v})\), and also
\begin{equation}
  \bm{\zeta} = (\bm{\zeta}_1, \dots , \bm{\zeta}_{n+1})^T\qquad \, 
  \bm{0} = (0, \dots , 0)^T\; .
\end{equation}
\end{subequations}
 \qed

@@

Soem future interesting directions. We might want to calculate the
solutions of the KP equation to see what they really mean. Perhaps
consider analitic and numerical calculations, since the calculation of
the determinants become complicated even for small values of \(n\).

>>>>

\title[Slavnov products, KP and BA functions]{
  More on the Slavnov products of spin chains,
  KP hierarchy tau and Baker-Akhiezer functions 
}

@@

\address{\noindent 
Institute of Exact Sciences, Department of Physics\\
Fluminense Federal University\\
Rua Des. Ellis Hermydio Figueira, 783, Aterrado\\
27213-145 Volta Redonda, RJ, Brazil
}

@@

\begin{abstract}
Connections between classical and quantum integrable systems are
analyzed from the viewpoint of the Slavnov products of Bethe
states. It is well known that, modulo model-dependent aspects, the
functional structure of the Slavnov products takes the form of
determinants. Using new results on the general structure of rational
and trigonometric models, we establish that, as long as the model
satisfies certain conditions, its Slavnov product is a tau function of
the KP hierarchy, thus generalizing some known results. We also show
that the Slavnov products can be expanded in terms of other tau
functions, and we prove that their homogeneous limit can be calculated
as a Wronskian of functions related to the eigenvalues of the transfer
matrices. Finally, we calculate the Baker–Akhiezer functions
associated with these Slavnov products, and show that, modulo a
multiplicative universal term, the Baker–Akhiezer functions have a
closed determinantal formula.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

One important ingredient in the Algebraic Bethe Ansatz is the scalar
product of Bethe states. This object plays a central role in the
analysis of quantum integrable systems. Many important results have
been collected on this topic, see~\cite{Korepin:1993kvr} and
references therein. In this text, we focus on a powerful and elegant
aspect of these products: scalars products in the algebraic Bethe
ansatz can be expressed in terms of determinants.

More specifically, we investigate a very special type of correlation
function in quantum integrable spin chains, the so called
\emph{Slavnov product}~\cite{Slavnov1989}. These objects can be seen
as building blocks of more sophisticated correlation functions and are
instrumental in the study of norms of Bethe wave
functions~\cite{Korepin:1993kvr}. But more importantly for the
purposes of the present work, Slavnov products seem to be a bridge
connecting quantum and classical integrable system. This is the
problem we explore in this work.

Connections between classical and quantum integrable systems are a
long-standing research area, and have been addressed from a wide range
of perspectives~\cite{Wu:1975mw, Its:1992bj, korepin2000the, Foda2009,
  Alexandrov:2011aa}.  In~\cite{Foda:2009zz}, the authors started the
analysis of the connections between the Slavnov products of the
Heisenberg XXZ spin chain and the Kadomtsev–Petviashvili (KP)
classical integrable hierarchy. Their analysis has been extended in
several different directions and applications, for
example~\cite{Wheeler:2010vmq, Foda:2010, Takasaki:2010qm,
  Foda:2012wn, Foda:2012wf}.  In~\cite{Araujo:2021ghu,
  Araujo:2024klz}. The author of the current paper also investigated
some aspects of this research problem. In particular, the relation
between integrable hierarchies and quantum integrable systems such as
the Q-Boson integrable system and the Temperley-Lieb open spin chains.

The main ingredients for the Slavnov product are the Bethe
wavefunctions themselves. As such, the specific details of these
objects depend on the model, symmetries and their representations, and
on the boundary conditions. Nevertheless, the general
\textbf{functional structure} of Slavnov products across radically
different models is basically the same. More specifically, modulo some
multiplicative factors, all these expressions take the form of
determinants, and it is this basic property that allows us to
manipulate these objects and prove that they are KP and/or Toda tau
functions.

Belliard and Slavnov~\cite{Belliard:2019bfz} have thoroughly
investigated this property for spin chains with rational and
trigonometric \(R\)-matrices, and basically answered ``\emph{why
scalar products in the Algebraic Bethe Ansatz have determinant
representations}''. Using their results, we now want to understand
``\emph{why and how (some) scalar products in the Algebraic Bethe
Ansatz are KP and Toda tau functions}''. We prove that, as long as the
integrable system satisfies the conditions established
in~\cite{Belliard:2019bfz} -- along with some additional restrictions
-- the Slavnov product is guaranteed to satisfy the KP integrable
hierarchy. Additionally, we discuss some immediate consequences of
this result.

In Section 2, we review the main aspects of the
work~\cite{Belliard:2019bfz}. This section also fixes some of the
notation we use in the remainder of this work and in future
investigations. Section 3 presents two new results. First, we rewrite
relevant formulas established in~\cite{Belliard:2019bfz} and prove
that these Slavnov products are KP tau functions expressed in terms of
alternant determinants. We also discuss some contrasts between this
formula and certain results that have previously appeared in matrix
theory investigations.

These tau functions are written in terms of two sets of
parameters. The first set is given by arbitrary complex numbers, and
the second set is given by the Bethe roots of the spin chain.  In
Section 3, we also find a basis for the Slavnov products, and this
basis is itself formed by tau functions. The results of Section 3
allow us to investigate the case where the complex parameters are
close to the Bethe roots.

Section 4 discusses the homogeneous limit of the Slavnov product.  In
this case, all complex parameters condense to a single value, and we
show that Slavnov products have a Wronskian expression written in
terms of functions related to the eigenvalues of the transfer matrix.
Section 5 presents some explicit examples for small spin chains, and
we see that even in those cases the expressions for the tau functions
become overwhelming very quickly.

Finally, in Section 6, we use the Japanese formula to investigate the
Baker–Akhiezer function associated with these tau functions. Using
known integral formulas for the tau functions, we write them in terms
of Miwa coordinates. The most important result in this section is the
explicit expressions for the Baker–Akhiezer functions. We discuss some
interesting consequences and open problems in Section 7.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Scalar products as determinants}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To ensure the paper is as self-contained as possible, this section
reviews the main arguments of the work of Belliard and
Slavnov~\cite{Belliard:2019bfz}. Let us start with a set of arbitrary
complex parameters \(\bm{u} = \{u_j\}_{j=1}^{n+1}\), and define
\(n+1\) sets \(\bm{u}_j=\bm{u}\setminus \{u_j\}\). In the Algebraic
Bethe Ansatz context~\cite{Korepin:1993kvr, Slavnov:2019aba}, we can
now define \(n+1\) off-shell Bethe vectors \(\ket{\Psi(\bm{u}_j)}\),
i.e., Bethe states where the algebraic Bethe equations are not imposed
on the parameters \(\bm{u}_j\). Additionally, we must also consider a
set of Bethe roots \(\bm{v} = \{ v_k \}_{k=1}^n\), that is, parameters
\(\bm{v}\) that satisfy the algebraic Bethe equations. Finally, we
define the on-shell Bethe vectors \(\ket{\Psi(\bm{v})}\).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Determinant representation}
The action of the transfer matrix \(\mathcal{T}(z)\), that is an
Hermitian operator, on the dual on-shell Bethe vector is given by
\begin{equation}
\label{eq:eigenvalue-bethe}
\bra{\Psi(\bm{v})}\mathcal{T}(z) = \Lambda(z; \bm{v}) \bra{\Psi(\bm{v})} \; , 
\end{equation}
where \(\Lambda(z; \bm{v}) \) is the transfer matrix eigenvalue.

We construct \(n+1\) functions obtained from products between the
on-shell and each off-shell Bethe states
\begin{equation}
  \bm{\zeta}_j(\bm{u}_j, \bm{v}) 
  = \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)}\; ,\qquad j =1, \dots, n+1\; .
\end{equation}
These partially on-shell scalar products are called \emph{Slavnov
products}.  Henceforth, the functional dependence of these functions
will be omitted, that is \(\bm{\zeta}_j\equiv \bm{\zeta}_j(\bm{u}_j,
\bm{v})\).

In this work we only consider integrable models where the action of
the transfer matrix on a generic off-shell bethe states can be
expanded as
\begin{subequations}
\begin{equation}
\label{eq:scalar-exp}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)}  = \sum_{k=1}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; ,
\end{equation}
where \(L_{jk}\) are coefficients, and the off-diagonal elements
\(L_{jk}\) with \(j\neq k\) contain the unwanted
terms~\cite{Slavnov:2019aba}. From the Algebraic Bethe Ansatz
construction, the non-diagonal terms must vanish when the Bethe
equations are satisfied, and this is basically the definition of the
algebraic bethe equations. Consequently, only the diagonal
coefficients \(L_{jj}\) survive, and we conclude that these terms must
be equal to the eigenvalue of the tranfer matrix.  Therefore
\begin{equation}
  \mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} =  \Lambda(u_j; \bm{u}_j) \ket{\Psi(\bm{v})} + 
  \sum_{\substack{k=1\\ k\neq j}}^{n+1} L_{jk} \ket{\Psi(\bm{u}_k)} \; . 
\end{equation}
\end{subequations}

This class of models includes many of the most familiar spin chains
with rational and trigonometric \(R\)-matrices -- including those
models discussed in the introduction. One important class of models
that does not fall into this classification is defined by
elliptic \(R\)-matrices, although some of these models still have slavnov
products with determinantal representations.

\begin{proposition}
The Slavnov products \(\bm{\zeta}_j\) satisfy a system of linear
equations~\cite{Slavnov:2019aba}.
\end{proposition}
\paragraph{\textbf{\emph{Proof}}} Let us first use that the
transfer matrix can act on the on-shell bra \(\bra{\Psi(\bm{v})}\)
on on the off-shell ket \(\ket{\Psi(\bm{u}_j)}\), then  
\begin{equation}
\label{eq:scalar-prod}
  \bra{\Psi(\bm{v})} \left(\mathcal{T}(u_j)\ket{\Psi(\bm{u}_j)} \right) =
  \left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)\ket{\Psi(\bm{u}_j)}
  \qquad j =1, \dots, n+1\; .
\end{equation}
The right-hand side of this equation can be simplified with the eigenvalue
expression~(\ref{eq:eigenvalue-bethe}), that is
\begin{equation}
  \left(\bra{\Psi(\bm{v})}\mathcal{T}(u_j)\right)\ket{\Psi(\bm{u}_j)}  =
\Lambda(u_j; \bm{v}) \bracket{\Psi(\bm{v})}{\Psi(\bm{u}_j)} = 
\Lambda(u_j; \bm{v}) \bm{\zeta}_j\; . 
\end{equation}

@@

Puttting all these facts together, we can write the
expression~(\ref{eq:scalar-prod}) as
\begin{equation}
  \sum_{k=1}^{n+1} L_{jk} \bm{\zeta}_k  = 
  \Lambda(u_j; \bar{v}) \bm{\zeta}_j\quad \Rightarrow \quad
  \sum_{k=1}^{n+1} M_{jk} \bm{\zeta}_k = 0 \; ,
\end{equation}
where we have defined the coefficients
\begin{equation}
  M_{jk} = L_{jk} - \delta_{jk}\Lambda(u_j; \bm{v})\; . 
\end{equation}
In simple terms, this expression shows that the Slavnov products
satisfy a linear system.  We can write this system in a matrix form as
\begin{equation}
\bm{M} \bm{\zeta} = 
  \left(\bm{L} - \bm{\Lambda} \right) \bm{\zeta} = \bm{0}\; ,
\end{equation}
where
\begin{subequations}
  \begin{equation}
    \bm{L} = 
 \begin{pmatrix}
   L_{1,1} & \cdots & L_{1,n+1}\\
   L_{2,1} & \cdots & L_{2,n+1}\\
   \vdots & \vdots & \vdots\\
   L_{n+1, 1} & \cdots & L_{n+1, n+1}\\
 \end{pmatrix}  \quad , \quad
    \bm{\Lambda} = \mathrm{diag}(\Lambda_1, \Lambda_2, \dots, \Lambda_{n+1})
\end{equation}
with \(\Lambda_j \equiv \Lambda(u_j; \bm{v})\), and
\begin{equation}
  \bm{\zeta} = (\bm{\zeta}_1, \dots , \bm{\zeta}_{n+1})^T\qquad \, 
  \bm{0} = (0, \dots , 0)^T\; .
\end{equation}
\end{subequations}
This completes the proof of their proposition. 
 \qed

@@

Some future interesting directions. We might want to calculate the
solutions of the KP equation to see what they really mean. Perhaps
consider analitic and numerical calculations, since the calculation of
the determinants become complicated even for small values of \(n\).
